On joint estimation of ancestral states and phylogenetic model parameters

### 2017-10-10 joint_inf.tex
a few ideas trying to work through general case
clarify and update plot

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

@@

    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \\

@@

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w')
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w').
\end{align*}
Since
$$
\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, we focus on the first likelihood.
Simplifying and suppressing some variables, this likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} + \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
\begin{align*}
f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
\end{align*}
implying $\ell'$ has two zeros according to the quadratic formula.
By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.

>>>>

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
% In Semple/Steel, there are some assumptions that are necessary for some of the results, called M1, M2 and M3.
% M1 and M2 seem definitely necessary, which pretty much just imply x and y can't be zero.
% M3 implies x and y can't be one?
% (das TODO: double check these; currently working from notes and memory)
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

\begin{align}
    \label{eq:farris_likelihood}

@@

    &=        p_{\emptyset}  \cdot\log(1+x'^2+y'^2+4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{3}          \cdot\log(1-x'^2+y'^2-x'^2y'^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{13}         \cdot\log(1+x'^2+y'^2-4x'y'w'+x'^2y'^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log(1-x'^2-y'^2+x'^2y'^2) \nonumber \\
    &\qquad + p_{123}        \cdot\log(1+x'^2-y'^2-x'^2y'^2) \nonumber \\
    &\qquad - \log(8) \nonumber \\
    &\qquad + p_{\emptyset}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{1}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{2}          \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{3}          \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \nonumber \\
    &\qquad + p_{12}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\
    &\qquad + p_{13}         \cdot\log((1+x')^2   (1-w')(1+y')^2) \nonumber \\
    &\qquad + p_{23}         \cdot\log((1+x')(1-x')(1+w')(1+y')(1-y')) \nonumber \\

@@

\end{align}

@@

First, assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
Recalling Table~\ref{tab:likelihoods}, the likelihood can take one of three forms corresponding to which ancestral state for $\siteSplit_j=\{1,3\}$ maximizes the likelihood for $(x',y',w')$.
If $x'$ and $y'$ are fixed, the likelihood in \eqref{eq:farris_likelihood} as a function of $w'$ is proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log(1+w') \\
    &\qquad + p_{13}\cdot\log(1-w')
\end{align*}
by collapsing like terms and ignoring terms with only $x'$ and $y'$.
Similarly, the other two possibilities for the likelihood are both proportional to
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log(1+w').
\end{align*}
Since
$$
\log(1+w') >= (1 - p_{13})\cdot\log(1+w') + p_{13}\cdot\log(1-w')
$$
with equality only when $w'=0$, we focus on the second likelihood.
Simplifying and suppressing some variables, this likelihood can be written
$$
\ell(w') = \alpha_1\log(\beta+\gamma w') + \alpha_2\log(\beta-\gamma w') + \log(1+w')
$$
where
$$
\alpha_1 := \alpha_1(x, y) = 1+x^2+y^2+4xy^2+x^2y^2,
$$
$$
\alpha_2 := \alpha_2(x, y) = 1+x^2+y^2-4xy^2+x^2y^2,
$$
$$
\beta := \beta(x, y) = 1+x^2+y^2+x^2y^2,
$$
$$
\gamma := \gamma(x, y) = 4xy.
$$
Since $x$ and $y$ fall between zero and one, we make use of the inequalities (excepting the cases of $x=y=0$ and $x=y=1$)
$$
\alpha_1 > \alpha_2
$$
and
$$
\beta > \gamma.
$$
% Not sure if I need to show this, but if x and y aren't 1 it's 1+x^2+y^2+x^2y^2 > 4xy => (1+x^2)(1+y^2) > 4xy and 1+x^2 > 2x, 1+y^2 > 2y qed
The derivative is
$$
\ell'(w') := \frac{d}{dw'} \ell(w') = \frac{\alpha_1 \gamma}{\beta+\gamma w'} - \frac{\alpha_2 \gamma}{\beta-\gamma w'} + \frac{1}{1+w'}.
$$
The inequality $\beta > \gamma$ implies that any inequality relating $\ell'$ and 0 is equivalent to the same inequality relating 0 and the quadratic function
\begin{align*}
f(w') &= w^2\cdot(-\gamma^2\alpha_1-\gamma^2\alpha_2-\gamma^2) \\
      &\qquad + w\cdot(\gamma\alpha_1\beta-\gamma^2\alpha_1-\gamma\alpha_2\beta-\gamma^2\alpha_2) \\
      &\qquad + (\gamma\alpha_1\beta-\gamma\alpha_2\beta+\beta^2), \\
\end{align*}
implying $\ell'$ has two zeros according to the quadratic formula.
By $\alpha_1 > \alpha_2$, $\ell'(0) > 0$ implies $\ell$ is increasing at $w'=0$.
We use this fact to show that, if the smaller of the zeros of $f(w')$ is greater than one, $\hat{w}' \equiv 1$.
Using the quadratic formula with
$$
a(x, y) = -\gamma^2 \alpha_1 - \gamma^2 \alpha_2 - \gamma^2,
$$
$$
b(x, y) = \gamma \alpha_1 \beta - \gamma^2\alpha_1 - \gamma \alpha_2 \beta - \gamma^2 \alpha_2,
$$
$$
c(x, y) = \gamma \alpha_1 \beta - \gamma \alpha_2 \beta + \beta^2,
$$
the smaller zero is
$$
\hat{w}'(x, y) = \frac{-b(x, y) - \sqrt{b(x, y)^2 - 4a(x, y)c(x, y)}}{2a(x, y)},
$$
which is a function of the generating parameters $x$ and $y$.
We plot the region where $\hat{w}'(x, y) \ge 1$ in Figure~\ref{fig:bl-inconsistency}.
